---
title: "Developmental analyses, part 2"
description: "How to deal with longitudinal brain development data"
author: "Jason Lerch"
date: "2023/11/01"
categories: [R, longitudinal]
draft: true
---

### A digression into working with hierarchical trees

Let's see which brain areas have the biggest sex effects at age 65. For that we have to fit these models across all bits in our brain hierarchy:

```{r}
#| eval: false
#| cache: true

library(splines)
library(broom.mixed)
library(knitr)

# this turns the volumes from the hierarchy into a list of nodes for easier processing
brainstructs <- Traverse(hvols)

# estimate the interaction model for every node
sexModels <- map(brainstructs, ~ lmer(roi ~ ns(age,3) * Sex + (1|subject_id), 
                                      gf %>% mutate(roi = .x$volumes)))

# estimate the sex effect at three ages for every model
ageEffects <- map(sexModels, ~ emmeans(.x, pairwise ~ Sex | age, at = list(age=c(7, 21, 65))))

# keep just the t-stats for now
ageEffectsT <- map_dfr(ageEffects, ~ tidy(.x$contrasts) %>% 
                         dplyr::select(age, statistic) %>% 
                         pivot_wider(names_from = age, values_from = statistic, names_prefix = "p"))

# now put the effects back onto the tree
hvols$Set(sexAtP7 = ageEffectsT$p7,
          sexAtP21 = ageEffectsT$p21,
          sexAtP65 = ageEffectsT$p65)

# print it
print(hvols, "sexAtP65")

# look at just the partly significant bits
hvols$Get("sexAtP65") %>% as.data.frame() %>% filter(abs(.) > 2) %>% arrange(desc(.))

```

So this works and has a decent bit of face validity given what we know about sex differences in the brain.

Let's put it onto the brain

```{r}
#| eval: false
library(MRIcrotome)
library(grid)

# let's display things symmetrically
hvolsSym <- Clone(hvols)
Prune(hvolsSym, function(x) !str_starts(x$name, "right") & !str_starts(x$name, "left"))

sliceSeries(nrow=6, ncol=1, begin=65, end=210) %>%
  anatomy(mritemplate, low=700, high=1400) %>%
  sliceSeries() %>% anatomy() %>%
  overlay(hanatToVolume(hvolsSym, mrilabels, "sexAtP21"), low=2, high=6, symmetric = T) %>%
  addtitle("P21") %>%
  sliceSeries() %>% anatomy() %>%
  overlay(hanatToVolume(hvolsSym, mrilabels, "sexAtP65"), low=2, high=6, symmetric = T) %>%
  addtitle("P65") %>%
  legend("t-statistics") %>%
  draw()
```

Let's test that for every brain structure

```{r}
#| eval: false
#| cache: true

simpleAgeModels <- map(brainstructs, ~ lmer(vol ~ ns(age, 3) + (1|subject_id), REML=F,
                                            gf %>% mutate(vol=.x$volumes)))
ageAndSexModels <- map(brainstructs, ~ lmer(vol ~ ns(age, 3) + Sex + (1|subject_id), REML=F,
                                            gf %>% mutate(vol=.x$volumes)))
ageAndSexIntModels <- map(brainstructs, ~ lmer(vol ~ ns(age, 3) * Sex + (1|subject_id), REML=F,
                                            gf %>% mutate(vol=.x$volumes)))

sexmodelComparison <- pmap(list(simpleAgeModels, ageAndSexModels, ageAndSexIntModels),
                           function(a,b,c) anova(a,b,c))

sexmodelComparisonP <- map_dfr(sexmodelComparison, ~ tidy(.x) %>%
                                 dplyr::select(term, p.value) %>%
                                 filter(!is.na(p.value)) %>%
                                 mutate(p.value=-log10(p.value)) %>%
                                 pivot_wider(names_from = term, values_from = p.value))

hvols$Set(sexInterceptP=sexmodelComparisonP$b,
          sexInteractP=sexmodelComparisonP$c)

ToDataFrameNetwork(hvols, "name", "sexInterceptP", "sexInteractP") %>%
  arrange(desc(sexInterceptP)) %>% 
  dplyr::select(-from, -to) %>%
  kable()
```

And display them on a brain

```{r}
#| eval: false
# let's display things symmetrically
hvolsSym <- Clone(hvols)
Prune(hvolsSym, function(x) !str_starts(x$name, "right") & !str_starts(x$name, "left"))

sliceSeries(nrow=6, ncol=1, begin=65, end=210) %>%
  anatomy(mritemplate, low=700, high=1400) %>%
  sliceSeries() %>% anatomy() %>%
  overlay(hanatToVolume(hvolsSym, mrilabels, "sexInterceptP"), low=1, high=3, symmetric = F) %>%
  addtitle("Sex Intercept") %>%
  sliceSeries() %>% anatomy() %>%
  overlay(hanatToVolume(hvolsSym, mrilabels, "sexInteractP"), low=1, high=3, symmetric = F) %>%
  addtitle("Sex Interaction") %>%
  legend("-log10 p value") %>%
  draw()
```

Again some high face validity results here, and now embedded tests of the "main effects"

There are bunch of warnings for singular fits, which will come up for the interaction models at times. Something to worry about for more detailed analyses.

