---
title: "MRIcrotome and MRIcroscope"
description: "Introducing the revised MRIcrotome"
author: "Jason Lerch"
date: "2025/08/16"
categories: [RMINC]
draft: true
engine: knitr
number-selections: false
---

I've been rewriting MRIcrotome, our package for visualizing brain slices, and it's time to introduce it to the world. But first, some history and background.

### Quick history of MRIcrotome

Early versions of RMINC included a way to display a slice series. While quite handy, it lacked flexibility. And so MRIcrotome was born, the portmanteau describing what it did well: describe a way to slice up 3D MRI scans to display in 2d. It was written from the ground up in [grid graphics](https://bookdown.org/rdpeng/RProgDA/the-grid-package.html), providing ultimate flexibility in creating visualizations.

That flexibility came at the cost of having to reinvent many wheels. This was particular around legends - the code for displaying a series of slices was quite simple compared to the code of displaying the colour bar. This came to a head when I wanted to implement code to display brain segmentation outlines on top of brain slice: clearly doable, but nicely done already within many mapping libraries (such as sf) that starting from scratch seemed silly.

So I started again, inspired primarily by vaguely remembered code from [Yohan Yee](http://yohanyee.ca), who had long ago shown examples of how to display beautiful brain slices in ggplot. Additional inspiration came from [ggseg](https://ggseg.github.io/ggseg/), a package that does a nice job of showing 2d renderings of brain surfaces and slices, but (as far as I can tell) focuses on pre-existing atlases over flexible display of arbitrary slices and atlases.

### Installation

The code still lives in its own github branch, and, full warning, the interface is most definitely liable to change. To get the current version you need the following:

```{r}
#| eval: false
devtools::install_github("Mouse-Imaging-Centre/MRIcrotome", ref="ggMRIcrotome")
```

You will also need the following packages installed:

- ggplot2
- sf
- terra
- tidyterra
- smoothr

For those of you on Oxford's clusters you'll find this development version of MRIcrotome already installed, so don't need to do anything.

### Data loading and modelling preliminaries

Before getting into visualizing let's first load some data and run some models so that we have something to work with. I'll work with the [Sex Chromosome Trisomy](https://pubmed.ncbi.nlm.nih.gov/23926958/) mouse model, also discussed [here](../sct-power-basic/index.qmd).

```{r}
#| output: false
# load the libraries
library(tidyverse)
library(RMINC)

# base dir where files are kept
basedir <- "/Users/jason/data/"
gf <- read_csv(paste0(basedir, "mbase-pairwise/SCT.csv")) %>%
    mutate(localJacobians = paste0(basedir, localJacobians))
```

Now let's load the segmentations

```{r}
#| cache: true

# nlin and label file
labelFile <- paste0(basedir, "mbase-pairwise/scanbase_second_level-nlin-3_labels.mnc")
nlinFile <- paste0(basedir, "mbase-pairwise/scanbase_second_level-nlin-3.mnc")

# definition files
defs <- "/Users/jason/data/atlases/Dorr_2008_Steadman_2013_Ullmann_2013_Richards_2011_Qiu_2016_Egan_2015_40micron/mappings/DSURQE_40micron_R_mapping.csv"
abi <- "/Users/jason/data/atlases/Allen_hierarchy_definitions.json"

# load the volumes - just integrating Jacobians here, for most projects
# you'd want to use the voted labels instead
allvols <- anatGetAll(gf$localJacobians, labelFile, defs=defs)

# make the atlas definitions hierarchical
hdefs <- makeMICeDefsHierachical(defs, abi)

# and add the volumes to the hierarchy
hvols <- addVolumesToHierarchy(hdefs, allvols)
```

And finally we can run some linear models - first at every voxel, then at every part of the segmentation. We'll model the data as Gonads (Testes vs Ovaries) plus X and Y chromosome dose.

```{r}
# load label and nlin volumes
labelVol <- mincArray(mincGetVolume(labelFile))
nlin <- mincArray(mincGetVolume(nlinFile))

# run a linear model on the relative jacobians
vs <- mincLm(localJacobians ~ G + X + Y, gf, mask=labelFile)

# run a linear model on the ROIs, covarying for brain volume
hlm <- hanatLm(~ G + X + Y, gf, hvols)
```

### Introducing MRIcrotome and MRIcroscope

The revised MRIcrotome package separates the display of slices into two steps: (1) MRIcrotome, slicing up the 3D brain into 2D slices, and (2) MRIcroscope (sorry ...) for taking those slices and displaying them using ggplot and friends on the back-end.

So let's slice up the brain. 

First we load the necessary libraries:

```{r}
#| output: false

library(MRIcrotome)
library(terra)
library(sf)
library(smoothr)
library(data.tree)
```

Then we slice and dice the brain

```{r}
#| output: false

sliceList <- MRIcrotome(nlin, labelVol, hvols, 
list(c(100,2), c(150,2), c(200,2), c(250,2), c(260, 2), c(300, 2), c(325,2), c(350,2), c(150,1)))
```

`MRIcrotome` takes four required arguments:

1. The anatomy volume.
2. The label volume.
3. The hierarchical label definitions.
4. A list of slice locations. Each slice is itself a vector consisting of the slice number and the dimension.

Let's take a look - we'll show the background anatomy, an overlay of the atlas, and the ROI based output for the gonads term.

```{r}
MRIcroscope(sliceList) %>%
    add_anatomy(low=500, high=1400) %>%
    add_roi_outline() %>%
    add_roi_overlay(data=hlm, column=tvalue.GM)
```

Or, one more example, let's overlay the voxel map of the same stats term instead:

